Reduced growth due to belowground sink limitation is not explained by reduced photosynthesis

### 2016-03-23 manuscript/manuscript_pve.Rmd
manuscript edits BM

\
<br/>
<br/>
<br/>
<br/>
One approach to examine the coordination of *A*~n~ and growth is to manipulate C source activity. Examples include elevated CO~2~ and partial defoliation experiments. Elevated CO~2~ has been shown to increase *A*~n~ [@drake1997more; @ainsworth2007response]; across four forest Free-Air CO~2~ Enrichment (FACE) experiments this increased photosynthetic uptake resulted in a stimulation of biomass production by 23 % on average [@norby2005forest]. Evidence from a wide range of elevated CO~2~ experiments, however, also reveals that even with an average photosynthetic enhancement of over 30 %, the biomass growth rate only increases by around 10 % [@kirschbaum2011does]. In partial defoliation experiments, increases in *A*~n~ of the remaining foliage are commonly shown, and are attributed to reduction in end-product inhibition [@iglesias2002regulation; @zhou2003changes; @handa2005test], enhanced biochemical activity [@ovaska1993b;@layne1995end], increased stomatal conductance [@layne1995end], enhanced leaf nutrient status [@turnbull2007increased] and regulatory sugar signaling [@eyles2013whole]. However, increases in *A*~n~ in defoliation experiments did not always produce increased growth due to reductions in meristem sink strength [@palacio2012fast], C limitation to mycorrhizal colonization [@markkola2004defoliation], or an overall decrease in whole plant C gain [@ovaska1993a]. These manipulations of C source activity demonstrate that changes in *A*~n~ do not always correspond with proportional responses of growth.

\
<br/>
<br/>
<br/>
<br/>
Alternatively, manipulation of plant tissue C sinks is often used to investigate the correlation between *A*~n~ and growth. This is because metabolic signaling networks, relaying information on C and N status of different tissues, can regulate photosynthetic activity [@paul2001sink]. If sink inhibition of *A*~n~ occurs, a close coordination between declines in *A*~n~ and growth should be expected. The occurrence of photosynthetic down regulation has been tested through fruit removal and phloem girdling. In these studies, down regulation of *A*~n~ was found to correlate with carbohydrate accumulation resulting from reduced tissue sink strength  [@iglesias2002regulation;@urban2007girdling; @haouari2013fruit]. However, reductions in *A*~n~ were also attributed to biochemical limitations prior to carbohydrate accumulation [@nebauer2011photosynthesis], irreversible photo-oxidative damage [@duan2008photosynthetic] and stomatal limitation [@li2005photosynthesis]. This lack of continuity among findings is not surprising, as we still know little about the balance between assimilation, storage and growth across temporal scales in plants [@smith2007coordination]. As these manipulations likely impact source as well as sink activity simultaneously, affect water transport, are very extreme, or are specific to the occurrence of large fruiting sinks, they tell us little about source-sink coordination in typical growing conditions for woody species.

\
<br/>
<br/>
<br/>
<br/>
An alternative experimental approach is to reduce belowground C sink strength in tree seedlings by manipulating rooting volume, by varying the container size [@arp1991effects; @nesmith1998effect; @poorter2012pot]. Possible advantages of this approach are that it allows a large range of treatment levels, can be easily compared to naturally planted seedlings and may mimic natural conditions as seedlings compete for space or reach bedrock. Seedlings undergo many physiological and morphological changes in response to rooting volume, including changes in biomass partitioning, *A*~n~, water relations, nutrient uptake and respiration [@nesmith1998effect;@poorter2012pot and references therein]. Inadequate rooting volume may decrease C sink strength by progressively restricting root growth in growing plants [@thomas1991root].  Container size studies frequently exhibit photosynthetic down-regulation, likely as a result of sink limitation [@arp1991effects; @mcconnaughay1991physical; @gunderson1994photosynthetic; @sage1994acclimation; @maina2002intra; @ronchi2006growth]. A meta-analysis by Poorter et al. [-@poorter2012pot] concluded that *A*~n~ is the process likely to be the strongest affected by pot size and may best explain the effects on biomass seen in the large number of studies where containers are used. This conclusion arises because plants grown in small containers are shown to accumulate leaf starch while having lower C exchange and assimilate export rates [@robbins1988effect]. However, evidence in support for a trade-off between C storage and growth in trees is, to date, inconclusive [@palacio2014does]. Based on these previous studies, using container size as a sink-strength manipulation can be used to empirically test the extent to which growth and *A*~n~ are coordinated.

\
<br/>
<br/>
<br/>
<br/>
This study utilizes a novel field design to investigate the coordination between growth and *A*~n~ in *Eucalyptus tereticornis* Sm. seedlings, by manipulating container size and thus rooting volume. Seedlings were maintained under well watered conditions in order to isolate the effect of restricted soil volume. We used freely-rooted seedlings as a control for the container size treatments. Empirical results were combined with a simple plant growth model to simulate seedling growth with a C mass balance approach, which was then compared to observed harvested seedling mass. The model used whole-plant C gain, scaled from instantaneous rates of leaf *A*~n~, to quantify seedling dry mass production over the 120 day experiment.

\
<br/>
<br/>
<br/>
<br/>
Our hypotheses were as follows: 
\
1). The manipulations of container size were expected to induce a belowground sink limitation compared to free seedlings. We hypothesized that declines in seedling growth would be largest in the smallest containers.

2). As the finite pool of rooting volume and soil nutrients will decline faster in trees growing in small containers, we expected reductions in partitioning to fine root mass relative to tree size with decreasing container size.

3). Reduced sink strength was expected to lead to accumulation of leaf non-structural carbohydrates, and a resulting down regulation of *A~n~*. We therefore expected a correlation between carbohydrate accumulation and photosynthetic capacity as a function of soil volume.

4). Last, observed seedling mass was expected to correspond to growth model mass predicted from a simple C balance model taking into account measured rates of photosynthesis.

@@

\
<br/>
<br/>
<br/>
<br/>
Forty-nine *Eucalyptus tereticornis* seedlings, 20 weeks old and approximately 40 cm tall in tube stock, were chosen from a single local Cumberland plain cohort. Six additional seedlings were harvested before planting to measure initial leaf area and dry mass of leaves, stems and roots. Previous container experiments have confirmed that species with tap roots (similar to *E. tereticornis*) use the center of the container as the medium for thick roots, leaving the periphery of the soil as the most active sites for fine root proliferation [@biran1980a; @biran1980b]. By using a species with tap root growth and manipulations of container length rather than width, we believed that a more realistic test of growth inhibition through constrained soil volume would be achieved.

\
<br/>
<br/>
<br/>
<br/>
Six container volumes were used ranging from 5 L to 35 L, with a 22.5 cm diameter, and lengths ranging from 15 to 100 cm. Containers were constructed of PVC pipe and were filled with local top soil (described above). Soil in each container was packed to achieve a target soil bulk density that matched local soil conditions of 1.7 g cm^-3^. An Imidacloprid (BAYER CropScience) insecticide tablet was planted 5 cm below the roots of each seedling. Containers were planted flush with the soil surface inside metal sleeves, designed to minimize excess air space between the container and outside soil while also allowing for container removal. This allowed for soil temperatures in containers to reflect conditions of naturally planted (free) seedlings. Each experimental block (n=7) contained a complete replicate set of six container volumes as well as one free seedling, with 1 m^2^ spacing. For each free seedling, used as the control, a 1 m^2^ subplot was excavated to the hard layer and replaced with the same soil used in each container. A border of root exclusion material was buried 0.25 m deep and extended 0.25 m above the ground surface around each subplot to exclude local vegetation, which was further kept out by periodic weeding.

\
<br/>
<br/>
<br/>
<br/>
Plants were watered weekly or when needed to maintain soil moisture at field capacity (13-15 %). Drain systems were built into each pot to prevent pooling of water throughout the experiment. Pooling of water could lead to an anaerobic environment around the root that could hinder the uptake of water through reduced root conductance [@poorter2009causes], an undesired experimental artifact. A collection compartment in the bottom of containers, containing gravel covered by root exclusion mesh, was used to collect excess water for 20, 25, and 35 l containers. Plastic tubing (6 mm diameter) was inset into the gravel layer and extended through the top of the container. A lysimeter pump was then used to suction excess water, through the tubing, as needed. For small containers (5, 10, and 15 L) a simple bottom plug was used to drain excess water from the gravel compartment. Each container was inspected after every watering and rainfall event to determine if pooling had occurred.

##Growth and morphology metrics
Seedlings were planted in summer (January 21^st^ 2013) and stem height, diameter at 15 cm and leaf count were measured weekly thereafter. Once the growth rate of individual plants had significantly declined a full biomass harvest was completed and the experiment ended (May 21^st^ 2013). Dry mass of leaves, stems, roots and total leaf area (LI-3100C Area Meter; LI-COR, Lincoln, NE, USA) were measured for each seedling. Mean individual leaf area for each harvested seedling was calculated by dividing total measured leaf area by total leaf count of only fully expanded leaves. Mean individual leaf area was then used to interpolate total seedling leaf area through time with weekly leaf counts. Root mass was collected by removing the roots system and passing soil from each container through a 1 mm sieve, washing, separating into fine and coarse roots (<2 mm and >2 mm diameter, respectively) and then drying to a constant mass. Roots of seedlings in containers were not considered pot bound, as clusters of roots along the soil-container interface were not observed. Roots from the free seedlings were collected by excavating each 1 m^2^ subplot to the hard layer and keeping only roots within the subplot. For each seedling, a sub-sample of washed fine roots was analyzed for root length using WhinoRhizo software (Regent Instruments Inc., Quebec, QC, Canada). Specific root length (SRL) is reported as the root length divided by the dry mass of each sub-sample (m g^-1^). Fine root length density (FRLD) for seedlings in containers is reported as the total fine root length divided by the volume of each container (m dm^-3^).

@@

\
<br/>
<br/>
<br/>
<br/>
Photosynthetic CO~2~ response (AC~i~) curves were measured at 25 &deg;C on a random subset of each container size (n=3) after new leaves were first produced (March 13-14^th^, 2013) and prior to the final harvest (May 14-15^th^, 2013). Each AC~i~ curve was started at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l^-1^ at 25 &deg;C at saturating light (above). From these curves the photosynthetic parameters, *J*~max~ and *V*~cmax~, were quantified using the biochemical model of @farquhar1980biochemical and fit with the 'plantecophys' package [@duursma2015] in R [@RDevelopmentCoreTeam2011] using default parameters.

\
<br/>
<br/>
<br/>
<br/>
Leaf dark respiration rates (*R*) were measured on each seedling on the same dates as AC~i~ curves. Freshly detached leaves were collected at least 1 hour after sundown and placed inside a conifer chamber attached to the Licor 6400. Measurements were taken at a reference [CO~2~] of 400 $\mu$l l^-1^ while leaf temperature was maintained at current ambient conditions. Reported values of *R* are standardized rates at 25 &deg;C using a *Q*~10~ value (1.86) developed for these seedlings in a separate experiment (Drake et al. unpublished). Leaf area and dry mass were recorded for each leaf during gas exchange campaigns.

##Leaf water potential
Predawn ($\Psi$~pd~) and midday ($\Psi$~l~) leaf water potentials were measured for each seedling using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA) on fully expanded leaves during the same time period as AC~i~ and *R*. Leaves were detached and immediately stored inside foil-covered bags before water potential measurements were performed. $\Psi$~pd~ was measured before sunrise and $\Psi$~l~ at midday 13:00-14:30 h. These measurements were used as a measure of static water stress on the seedlings [@sellin1999does] and to ensure that the bulk soil water availability was high enough for plants to avoid water stress as they became larger and roots filled the soil volume.

@@

\
<br/>
<br/>
<br/>
<br/>
Leaf total non-structural carbohydrate (TNC) concentration was analyzed on dried and milled leaf samples using a total starch assay kit (Megazyme International, Wicklow, Ireland) and includes the starch  and soluble sugar concentrations (mg g^-1^). Starch was quantified using a thermostable $\alpha$-amylase and amyloglucosidase assay [@McCleary_starch] and soluble sugars were determined following the anthrone method [@ebell1969variation]. Complete methods of the TNC assay are described in @mitchell2013drought. TNC-free specific leaf area (SLA~f~, m^2^ kg^-1^) for leaves sampled during gas exchange campaigns, was calculated by first subtracting the TNC content from individual dry leaf mass before dividing leaf area by leaf mass or the leaf N content. Similarly, TNC-free leaf N (N~f~, %) was calculated on all gas exchange leaves from leaf mass without TNC content.

@@

\
<br/>
<br/>
<br/>
<br/>
*C*~day~ was predicted by using a coupled photosynthesis - stomatal conductance model [@farquhar1980biochemical;@medlyn2011reconciling] with the 'plantecophys' package [@duursma2015] in R with the mean photosynthetic parameters (*J*~max~, *V*~cmax~, *R* and *g*~1~) for each treatment and meteorological data from an onsite weather station. *J*~max~ and *V*~cmax~ were estimated from AC~i~ curves (explained above), *R* was empirically measured and the *g*~1~ parameter was generated by fitting the optimal stomatal conductance model from [@medlyn2011reconciling] to observed *g*~s~ values. Methods of the coupled leaf gas exchange model are also described in Duursma et al [-@duursma2014peaked]. Combined with the meteorological parameters; PPFD, air temperature, and relative humidity, at 15 min intervals, leaf *A*~n~ rates ($\mu$mol CO~2~ m^-2^ s^-1^) were then predicted for each soil volume treatment. *C*~day~ was calculated by converting predicted rates to mass C gain over 15 min time steps (g m^-2^) and then summed for 24 h. This resulted in 120 unique values of *C*~day~ for each soil volume treatment, one value for each day of the experiment. Thus, each daily time step for model runs included a value of *C*~day~ that represented both treatment specific photosynthetic parameters and meteorological constraints across the duration of the experiment.

\
<br/>
<br/>
<br/>
<br/>
It was further necessary to calculate a self-shading parameter ($\sigma$~s~) when scaling leaf *A*~n~ with total plant leaf area. This was accomplished by utilizing 61 previously digitized *Eucalyptus* seedlings, covering 5 total species which include *E. tereticornis*, from Duursma et al. [-@duursma2012light] to run in 'YplantQMC' package [@YplantQMC] in R to build a 3D plant structure based on digitized metrics of plant allometry and crown structure. Using the same treatment specific physiological parameters listed above, 'YplantQMC' outputs total *A*~n~, using total leaf area, for seedlings assuming self-shading as well as for a full sun large horizontal leaf. The ratio of total *A*~n~ with self-shading to horizontal leaf was then used to calculate $\sigma$~s~ for each of the 61 digitized seedlings, independently for each of the seven soil volume treatments. Next, the linear relationship between $\sigma$~s~ and total leaf area was determined across digitized seedlings, within each treatment. For the growth model, $\sigma$~s~ was then predicted for each daily time step using the previous days cumulative leaf area and this value  was then applied to C~day,i~. All default parameters used in model simulations are reported in Table `r pvetabS`.

\
<br/>
<br/>
<br/>
<br/>
We then used this model to test the hypothesis that the effects of belowground sink limitation on rates of leaf *A*~n~ were sufficient to accurately predict overall seedling biomass production after 120 days. Each model run used changes in *A*~n~ and leaf mass fraction (LMF), with values of stem and root respiration rates, to generate total seedling mass and leaf area after 120 days.  Cumulative net leaf C gain for each treatment was equal to the sum of each value of C~day,i~ over 120 days and final seedling C was assumed to equal half of the final dry mass for both modeled and observed seedlings.

\
<br/>
<br/>
<br/>
<br/>
To test the hypothesis, a default model with mean photosynthetic parameters from the free seedlings was optimized to estimate a LMF which correctly predicted both the leaf mass and total biomass of the harvested free seedling (*M*~free~). This optimized LMF was then used to constrain model runs with treatment-specific *C*~day~, while keeping tissue respiration and turnover parameters constant, to determine if changes in leaf *A*~n~ alone could predict biomass of seedlings in containers (*M*~pots~). Next, model sensitivity to different C allocation scenarios, including observed treatment-specific LMF and up-regulation of non-leaf tissue respiration by 50 % of default values (*M*~S1~ and *M*~S2~, respectively), was used to test for improvement of model biomass predictions compared to measured harvest biomass. For all cases, seedling biomass production was compared between model output and harvested seedlings with treatment-specific mean *C*~day~ by first scaling values to the free seedling control.

@@

\
<br/>
<br/>
<br/>
<br/>
Total seedling biomass at harvest was significantly different across container volumes (*P* < 0.001) and between container treatments and free seedlings (*P* < 0.001, Table `r pvetab1`). On average, harvested biomass of free seedlings was 84% higher than that of seedlings in containers. Plant biomass was positively correlated with total leaf area across all treatments (*R*^2^ = 0.97, *P* < 0.001). Differences in biomass partitioning to leaves, stems and roots were not different across soil volumes after variation in seedling biomass across treatments was factored in the analysis (Figure `r allocfig`a,b). Across all treatments, the final harvest root:shoot biomass ratio was conserved in these seedlings which exhibited a slightly higher shoot than root mass ($\overline{x}$ = 0.904, 95% CI = [0.846,1.119]) and the ratio of leaf to fine root mass was also not different (Figure `r allocfig`c).

\
<br/>
<br/>
<br/>
<br/>
Overall, SRL was higher in seedlings in containers compared to free seedlings but only in some of the container size treatments (Table `r pvetab2`, *P* = 0.009). Fine root length density was significantly higher in the two smallest container sizes and was the lowest in the largest container size (Table `r pvetab2`, *P* < 0.001). Over the duration of the experiment SLA~f~ was higher in free seedlings, but was not different across container sizes (Table `r pvetab1`, *P* < 0.001) and this pattern was evident beginning in the first gas exchange measurement campaign (*P* < 0.001).

@@

\
<br/>
<br/>
<br/>
<br/>
Both *J*~max~ and *V*~cmax~ were significantly higher in free seedlings (30 % and 26 %, respectively) than container-grown seedlings with little variation between container volume treatments (Table `r pvetab3`). Leaf dark respiration rates were not significantly different across soil volumes (Table `r pvetab3`). The *g*~1~ parameter, generated for each seedling from the Medlyn et al [-@medlyn2011reconciling] optimal stomatal conductance model, was lowest in the free seedling treatment and was marginally different across soil volume treatments (Table `r pvetab3`).

>>>>

\
One approach to examine the coordination of *A*~n~ and growth is to manipulate C source activity. Examples include elevated CO~2~ and partial defoliation experiments. Elevated CO~2~ has been shown to increase *A*~n~ [@drake1997more; @ainsworth2007response]; across four forest Free-Air CO~2~ Enrichment (FACE) experiments this increased photosynthetic uptake resulted in a stimulation of biomass production by 23 % on average [@norby2005forest]. Evidence from a wide range of elevated CO~2~ experiments, however, also reveals that even with an average photosynthetic enhancement of over 30 %, the biomass growth rate only increases by around 10 % [@kirschbaum2011does]. In partial defoliation experiments, increases in *A*~n~ of the remaining foliage are commonly shown, and are attributed to reduction in end-product inhibition [@iglesias2002regulation; @zhou2003changes; @handa2005test], enhanced biochemical activity [@ovaska1993b;@layne1995end], increased stomatal conductance [@layne1995end], enhanced leaf nutrient status [@turnbull2007increased] and regulatory sugar signaling [@eyles2013whole]. Similar to elevated CO~2~ experiments, these increases in *A*~n~ in defoliation experiments did not always produce increased growth due to reductions in meristem sink strength [@palacio2012fast], C limitation to mycorrhizal colonization [@markkola2004defoliation], or an overall decrease in whole plant C gain [@ovaska1993a]. These manipulations of C source activity demonstrate that changes in *A*~n~ do not always correspond with proportional responses of growth.

\
Alternatively, manipulation of plant tissue C sinks is often used to investigate the correlation between *A*~n~ and growth. Plant metabolic signalling networks have the ability to regulate photosythetic activity, according to carbon status of different plant tissues [@paul2001sink]. If sink inhibition of *A*~n~ occurs, a close coordination between declines in *A*~n~ and growth should be expected. The occurrence of photosynthetic down regulation has been tested through fruit removal and phloem girdling. In these studies, down regulation of *A*~n~ was found to correlate with carbohydrate accumulation resulting from reduced tissue sink strength  [@iglesias2002regulation;@urban2007girdling; @haouari2013fruit]. However, reductions in *A*~n~ were also attributed to biochemical limitations prior to carbohydrate accumulation [@nebauer2011photosynthesis], irreversible photo-oxidative damage [@duan2008photosynthetic] and stomatal limitation [@li2005photosynthesis]. This lack of continuity among findings is not surprising, as we still know little about the balance between assimilation, storage and growth across temporal scales in plants [@smith2007coordination]. As these manipulations likely impact source as well as sink activity simultaneously, affect water transport, are very extreme, or are specific to the occurrence of large fruiting sinks, they tell us little about source-sink coordination in typical growing conditions for woody species.

\
An alternative experimental approach is to reduce belowground C sink strength in tree seedlings by manipulating rooting volume, by varying the container size [@arp1991effects; @nesmith1998effect; @poorter2012pot]. Inadequate rooting volume can decrease C sink strength by progressively restricting root growth in growing plants [@thomas1991root]. Container size studies frequently exhibit photosynthetic down-regulation, likely as a result of sink limitation [@arp1991effects; @mcconnaughay1991physical; @gunderson1994photosynthetic; @sage1994acclimation; @maina2002intra; @ronchi2006growth]. A meta-analysis by Poorter et al. [-@poorter2012pot] concluded that *A*~n~ is the process likely to be the strongest affected by pot size and may best explain the effects on biomass seen in the large number of studies where containers are used. This conclusion arises because plants grown in small containers are shown to accumulate leaf starch while having lower C exchange and assimilate export rates [@robbins1988effect].

However, evidence in support for a trade-off between C storage and growth in trees is, to date, inconclusive [@palacio2014does].  Whether or not down regulation of *A*~n~ and reductions of growth are quantitavely similiar has rarely been tested.

Based on these previous studies, using a range of container sizes as a sink-strength manipulation can be used to empirically test the extent to which growth and *A*~n~ are coordinated.

\
This study utilizes a novel field design to investigate the coordination between growth and *A*~n~ in *Eucalyptus tereticornis* Sm. seedlings, by manipulating container size and thus rooting volume. Seedlings were maintained under well watered conditions in order to isolate the effect of restricted soil volume. We used freely-rooted seedlings as a control for the container size treatments. We tracked leaf-level gas exchange, leaf carbohydrate accumulation and seedling allometry over the course of 4 months. Empirical results were combined with a simple plant growth model to simulate seedling growth with a C mass balance approach, which was then compared to observed harvested seedling mass. The model used whole-plant C gain, scaled from instantaneous rates of leaf *A*~n~, to quantify seedling dry mass production over the 120 day experiment.

\
Manipulations of container size were expected to induce a belowground sink limitation compared to free seedlings, resulting in growth reductions in the smallest containers. Specifically, reductions in biomass partitioning to root mass, relative to tree size, were predicted to occur with decreasing rooting volume and soil nutrients in container treatments. We also expected a correlation between carbohydrate accumulation and photosynthetic capacity as a function of soil volume, as reduced sink strength would lead to accumulation of leaf non-structural carbohydrates and down regulation of *A~n~*. Taking into account measured rates of photosynthesis, we expected biomass predictions from the growth model to correspond to observed seedling mass

@@

\
Forty-nine *Eucalyptus tereticornis* seedlings, 20 weeks old and approximately 40 cm tall in tube stock, were chosen from a single local Cumberland plain cohort. Six additional seedlings were harvested before planting to measure initial leaf area and dry mass of leaves, stems and roots. Previous container experiments have confirmed that species with tap roots (similar to *E. tereticornis*) use the center of the container as the medium for thick roots, leaving the periphery of the soil as the most active sites for fine root proliferation [@biran1980a; @biran1980b]. By using a species with tap root growth and manipulations of container length rather than width, we believed that a more realistic test of growth inhibition through constrained soil volume would be achieved.

\
Six container volumes were used ranging from 5 L to 35 L, with a 22.5 cm diameter, and lengths ranging from 15 to 100 cm. Containers were constructed of PVC pipe and were filled with local top soil (described above). Soil in each container was packed to achieve a target soil bulk density that matched local soil conditions of 1.7 g cm^-3^. An Imidacloprid (BAYER CropScience) insecticide tablet was planted 5 cm below the roots of each seedling. Containers were planted flush with the soil surface inside metal sleeves, designed to minimize excess air space between the container and outside soil while also allowing for container removal. This allowed for soil temperatures in containers to reflect conditions of naturally planted (free) seedlings. Each experimental block (n=7) contained a complete replicate set of six container volumes as well as one free seedling, with 1 m^2^ spacing. For each free seedling, used as the control, a 1 m^2^ subplot was excavated to the hard layer and replaced with the same soil used in each container. A border of root exclusion material was buried 0.25 m deep and extended 0.25 m above the ground surface around each subplot to exclude local vegetation, which was further kept out by periodic weeding.

\
Plants were watered weekly or when needed to maintain soil moisture at field capacity (13-15 %). Drain systems were built into each pot to prevent pooling of water throughout the experiment. Pooling of water could lead to an anaerobic environment around the root that could hinder the uptake of water through reduced root conductance [@poorter2009causes], an undesired experimental artifact. A collection compartment in the bottom of containers, containing gravel covered by root exclusion mesh, was used to collect excess water for 20, 25, and 35 l containers. Plastic tubing (6 mm diameter) was inset into the gravel layer and extended through the top of the container. A lysimeter pump was then used to suction excess water, through the tubing, as needed. For small containers (5, 10, and 15 L) a simple bottom plug was used to drain excess water from the gravel compartment. Each container was inspected after every watering and rainfall event to determine if pooling had occurred.

##Growth and morphology metrics
Seedlings were planted in summer (January 21^st^ 2013) and stem height, diameter at 15 cm and leaf count were measured weekly thereafter. After four months, a full biomass harvest was completed and the experiment ended (May 21^st^ 2013). Dry mass of leaves, stems, roots and total leaf area (LI-3100C Area Meter; LI-COR, Lincoln, NE, USA) were measured for each seedling. Mean individual leaf area for each harvested seedling was calculated by dividing total measured leaf area by total leaf count of only fully expanded leaves. Mean individual leaf area was then used to interpolate total seedling leaf area through time with weekly leaf counts. Root mass was collected by removing the roots system and passing soil from each container through a 1 mm sieve, washing, separating into fine and coarse roots (<2 mm and >2 mm diameter, respectively) and then drying to a constant mass. Roots of seedlings in containers were not considered pot bound, as clusters of roots along the soil-container interface were not observed. Roots from the free seedlings were collected by excavating each 1 m^2^ subplot to the hard layer and keeping only roots within the subplot. For each seedling, a sub-sample of washed fine roots was analyzed for root length using WhinoRhizo software (Regent Instruments Inc., Quebec, QC, Canada). Specific root length (SRL) is reported as the root length divided by the dry mass of each sub-sample (m g^-1^). Fine root length density (FRLD) for seedlings in containers is reported as the total fine root length divided by the volume of each container (m dm^-3^).

@@

\
Photosynthetic CO~2~ response (AC~i~) curves were measured at 25 &deg;C on a random subset of each container size (n=3) after new leaves were first produced (March 13-14^th^, 2013) and prior to the final harvest (May 14-15^th^, 2013). Each AC~i~ curve was started at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l^-1^ at 25 &deg;C at saturating light. From these curves the photosynthetic parameters, *J*~max~ and *V*~cmax~, were quantified using the biochemical model of @farquhar1980biochemical and fit with the 'plantecophys' package [@duursma2015] in R [@RDevelopmentCoreTeam2011].

\
Leaf dark respiration rates (*R*) were measured on each seedling on the same dates as AC~i~ curves. Freshly detached leaves were collected at least 1 hour after sundown and placed inside a conifer chamber attached to the Licor 6400. Measurements were taken at a reference [CO~2~] of 400 $\mu$l l^-1^ while leaf temperature was maintained at current ambient conditions. Reported values of *R* are standardized rates at 25 &deg;C using a *Q*~10~ value (1.86) developed for these seedlings in a separate experiment (Drake et al. unpublished). Leaf area and dry mass were recorded for each leaf during gas exchange campaigns.

##Leaf water potential
Predawn ($\Psi$~pd~) and midday ($\Psi$~l~) leaf water potentials were measured for each seedling using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA) on fully expanded leaves during the same time period as AC~i~ and *R*. Leaves were detached and immediately stored inside foil-covered bags before water potential measurements were performed. $\Psi$~pd~ was measured before sunrise and $\Psi$~l~ at 13:00 h (&pm30 m). These measurements were used as a measure of static water stress on the seedlings [@sellin1999does] and to ensure that the bulk soil water availability was high enough for plants to avoid water stress as they became larger and roots filled the soil volume.

@@

\
Leaf total non-structural carbohydrate (TNC) concentration was analyzed on dried and milled leaf samples using a total starch assay kit (Megazyme International, Wicklow, Ireland) and includes the starch  and soluble sugar concentrations (mg g^-1^). Starch was quantified using a thermostable $\alpha$-amylase and amyloglucosidase assay [@McCleary_starch] and soluble sugars were determined following the anthrone method [@ebell1969variation]. Complete methods of the TNC assay are described in @mitchell2013drought. TNC-free specific leaf area (SLA~f~, m^2^ kg^-1^) for leaves sampled during gas exchange campaigns, was calculated by first subtracting the TNC content from individual dry leaf mass before dividing leaf area by leaf mass or the leaf N content. Similarly, TNC-free leaf N (N~f~, %) was calculated on all gas exchange leaves from leaf mass without TNC content.

@@

\
*C*~day~ was predicted by using a coupled photosynthesis - stomatal conductance model [@farquhar1980biochemical;@medlyn2011reconciling] with the 'plantecophys' package [@duursma2015] in R with the mean photosynthetic parameters (*J*~max~, *V*~cmax~, *R* and *g*~1~) for each treatment and meteorological data from an onsite weather station. Values of *J*~max~ and *V*~cmax~ are mean estimates from AC~i~ curves over two measurement periods (explained above), *R* was empirically measured and the *g*~1~ parameter was generated by fitting the optimal stomatal conductance model from [@medlyn2011reconciling] to observed *g*~s~ values. Methods of the coupled leaf gas exchange model are also described in Duursma et al [-@duursma2014peaked]. Combined with the meteorological parameters; PPFD, air temperature, and relative humidity, at 15 min intervals, leaf *A*~n~ rates ($\mu$mol CO~2~ m^-2^ s^-1^) were then predicted for each soil volume treatment. *C*~day~ was calculated by converting predicted rates to mass C gain over 15 min time steps (g m^-2^) and then summed for 24 h. This resulted in 120 unique values of *C*~day~ for each soil volume treatment, one value for each day of the experiment. Thus, each daily time step for model runs included a value of *C*~day~ that represented both treatment specific photosynthetic parameters and meteorological constraints across the duration of the experiment.

\
It was further necessary to calculate a self-shading parameter ($\sigma$~s~) when scaling leaf *A*~n~ with total plant leaf area. This was accomplished by utilizing 61 previously digitized *Eucalyptus* seedlings, covering 5 total species which include *E. tereticornis*, from Duursma et al. [-@duursma2012light] to run in 'YplantQMC' package [@YplantQMC] in R to build a 3D plant structure based on digitized metrics of plant allometry and crown structure. Using the same treatment specific physiological parameters listed above, 'YplantQMC' outputs total *A*~n~, using total leaf area, for seedlings assuming self-shading as well as for a full sun large horizontal leaf. The ratio of total *A*~n~ with self-shading to horizontal leaf was then used to calculate $\sigma$~s~ for each of the 61 digitized seedlings, independently for each of the seven soil volume treatments. Next, the linear relationship between $\sigma$~s~ and total leaf area was determined across digitized seedlings, within each treatment. For the growth model, $\sigma$~s~ was then predicted for each daily time step using the previous days cumulative leaf area and this value  was then applied to C~day,i~. All default parameters used in model simulations are reported in Table `r pvetabS`.

\
We then used this model to test the hypothesis that the effects of belowground sink limitation on rates of leaf *A*~n~ were sufficient to accurately predict overall seedling biomass production after 120 days. Each model run used changes in *A*~n~ and leaf mass fraction (LMF), with values of stem and root respiration rates, to generate total seedling mass and leaf area after 120 days.  Cumulative net leaf C gain for each treatment was equal to the sum of each value of C~day,i~ over 120 days and final seedling C was assumed to equal half of the final dry mass for both modeled and observed seedlings.

\
To test the hypothesis, a default model with mean photosynthetic parameters from the free seedlings was optimized to estimate a LMF which correctly predicted both the leaf mass and total biomass of the harvested free seedling (*M*~free~). This optimized LMF was then used to constrain model runs with treatment-specific *C*~day~, while keeping tissue respiration and turnover parameters constant, to determine if changes in leaf *A*~n~ alone could predict biomass of seedlings in containers (*M*~pots~). Next, model sensitivity to different C allocation scenarios, including observed treatment-specific LMF and up-regulation of non-leaf tissue respiration by 50 % of default values (*M*~S1~ and *M*~S2~, respectively), was used to test for improvement of model biomass predictions compared to measured harvest biomass. For all cases, seedling biomass production was compared between model output and harvested seedlings with treatment-specific mean *C*~day~ by first scaling values to the free seedling control.

@@

\
Total seedling biomass at harvest was significantly different across container volumes (*P* < 0.001) and between container treatments and free seedlings (*P* < 0.001, Table `r pvetab1`). On average, harvested biomass of free seedlings was 84% higher than that of seedlings in containers. Plant biomass was positively correlated with total leaf area across all treatments (*R*^2^ = 0.97, *P* < 0.001). Differences in biomass partitioning to leaves, stems and roots were not different across soil volumes after variation in seedling biomass across treatments was factored in the analysis (Figure `r allocfig`a,b). Across all treatments, the final harvest root:shoot biomass ratio and  was conserved in these seedlings which exhibited a slightly higher shoot than root mass ($\overline{x}$ = 0.904, 95% CI = [0.846,1.119]) and the ratio of leaf to fine root mass was also not different (Figure `r allocfig`c).

\
Overall, SRL was higher in seedlings in containers compared to free seedlings but only in some of the container size treatments (Table `r pvetab2`, *P* = 0.009). Fine root length density was significantly higher in the two smallest container sizes and was the lowest in the largest container size (Table `r pvetab2`, *P* < 0.001). Over the duration of the experiment SLA~f~ was higher in free seedlings, but was not different across container sizes (Table `r pvetab1`, *P* < 0.001) and this pattern was evident beginning in the first gas exchange measurement campaign (*P* < 0.001).

@@

\
Both *J*~max~ and *V*~cmax~ were significantly higher in free seedlings (30 % and 26 %, respectively) than container-grown seedlings with little variation between container volume treatments (Table `r pvetab3`). Leaf dark respiration rates were not significantly different across soil volumes (Table `r pvetab3`). The *g*~1~ parameter, generated for each seedling from the Medlyn et al [-@medlyn2011reconciling] optimal stomatal conductance model, was lowest in the free seedling treatment and was marginally different across soil volume treatments (Table `r pvetab3`).

\
Neither $\Psi$~pd~ nor $\Psi$~l~ were different across treatments, with mean values of -0.27 and -1.2 MPa across all seedlings, respectively. Although *g*~s~ in free seedlings was generally higher than those in containers (Table `r pvetab3`, *P* < 0.001), the mean rates for all seedlings were high at 0.37 mol m^-2^ s^-1^ and did not change throughout the course of the experiment. Additionally, leaf $\delta$^13^C at final harvest was not different across treatments (Table `r pvetab1`). Combined, these indices provide strong evidence that water stress was not apparent on these seedlings throughout the experiment. Soil N at harvest was not different across soil volumes ($\bar{x}$ = 4.5 %), with minimal decreases from pre-planting value ($\bar{x}$ = 4.9 %). This similarity indicates that nutrient leaching from free seedlings or from draining of containers following natural rainfall events did not differ between treatments.
